---
title: "A prediction model to predict activity quality from activity monitors."
output:
  html_document:
    keep_md: yes
  pdf_document: default
---
Author: Rainer-Anton Englisch

## Executive summary

"Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset)."

Simply put: We want to build a prediction model to predict activity quality from activity monitors. 

## Predictor Analysis and Reduction 
Before letting caret to create a prediction model we will try to reduce the number of predictors in order to speed up the creation of the prediction model. First we load the training data set and remove the column "X"" to avoid overfitting by observation number.

```{r loadCSV, cache=TRUE}
set.seed(1313)
pml_training = read.csv("pml-training.csv")
pml_training = pml_training[,-1]
dim(pml_training)
```

Next we split the original training in a new training and a test data. The test data will be used later for estimating the out of sample error.

The activity quality of an observation is classified by the factor variable classe which we store in seperate variables for later use for training and prediction.

```{r createDataPartitions,message=FALSE,cache=TRUE,dependson='loadCSV'}
library(caret)
inTrain <- createDataPartition(pml_training$classe,p=0.75, list=FALSE)
training <- pml_training[inTrain,]
trainclasse <- training$classe
testing <- pml_training[-inTrain,]
testingclasse <- testing$classe
```

Next we throw out all predictors that have near zero variability because we assume
that these predictors have minimal influence on the prediction.

```{r removeNearZeroVarPredictors, cache=TRUE,dependson='removeNAsPredictors'}
nZV <- nearZeroVar(training,saveMetrics=TRUE)
# extract the predictor names which have near zero variability
nzvcolnames <- rownames(nZV[nZV$nzv==TRUE,])
# compute the index of these predictor  names in the training data frame
nzvcolindex <- which(names(training) %in% nzvcolnames)
# remove these predictors
training <- training[,-nzvcolindex]
# print these predictors
nzvcolnames
```
We removed `r length(nzvcolnames)` predictors in the training data frame.

Next we want to throw out predictors that have a great linear correlation.

```{r createCorrelationMatrix, cache=TRUE,dependson='removeNearZeroVarPredictors'}
#As the correlation matix can only be computed for numeric variables we need to identify numeric variables
colsnumeric <- sapply(training, is.numeric)

#compute the correlation matrix
cortraining <- cor(training[,colsnumeric],use="pairwise.complete.obs")
```

Within the computed correlation matrix we select all predictors that have a high correlation. Let us define a high corelation as a value equal or greater than 0.7. Thus let's find these predictors and remove them from the training data set.
```{r removeHighlyCorrelatedPredictors, cache=TRUE,dependson='createCorrelationMatrix'}
# retrieve variables that have a correlation greater or equal to 0.7
highlyCor <- findCorrelation(cortraining, 0.70,verbose=FALSE)
highlyCorcolnames <- colnames(training)[highlyCor]
# remove the highly correlated predictors from the training data frame
training <- training[,-highlyCor]
# print the removed predictors
highlyCorcolnames
```
We removed `r length(highlyCor)` highly correlated predictors.

Now we want to quickly create  a small prediction model in order to
query the importance of the variables for the prediction model.
Let's fit a model, print the important predictors and keep these predictors in the training data set.
```{r keepImportantPredictors, cache=TRUE,dependson='removeHighlyCorrelatedPredictors'}
modFit <- train(classe ~.,data=training,method="rpart")
# print summary of the model fit
modFit
# retrieve the variable importance list
importance <- varImp(modFit, scale=FALSE)
importance <- importance[[1]]
# bind the rownames as columns 
importance <- cbind(rownames(importance),importance )
# retrieve column names which have importance greater zero
impcolnames <- importance[importance$Overall>0.0,1]
# print important variables
impcolnames
# retrieve indexes of column names
impcolindex <- which(names(training) %in% impcolnames)
# keep important columns in training data frame
training <- training[,impcolindex]
```
Now that we have reduced the predictors significantly from 160 to `r length(impcolnames)` predictors we will fit a more complex machine learning algorithm based on random forest. Additionally we will use cross validation to minimize in order to get a more realistic out of sample error rate.

```{r trainFinalModel, cache=TRUE,dependson='keepImportantPredictors'}
fitControl <- trainControl(method = "repeatedcv",number = 10, repeats = 3)
preObj <- preProcess(training,method=c("knnImpute"))
trainingImputed <- predict(preObj,newdata=training)
modFit <- train(trainclasse ~.,data=trainingImputed,
                method="rf"
                #method="rpart"
              ,trControl = fitControl
              )
# print summary of model
print(modFit)
```

Here let's print a summary of our model fit.
```{r}
# make predictions
predictions <- predict(modFit, trainingImputed)
# summarize results
confusionMatrix(predictions, trainclasse)
print(modFit$finalModel)
```


Now let us compare the estimated OOB sample error versus the out of sample error. Therefore let's compute the out of sample error by predicting based on our model fit and the testing data set.
```{r preditTestingDataFrame, dependson='trainFinalModel'}
# subselect the predictors used for training
traincolnames <- colnames(training)
traincolindex <- which(names(testing) %in% traincolnames)
testing <- testing[,traincolindex]
preObj <- preProcess(testing,method=c("knnImpute"))
testingImputed <- predict(preObj,newdata=testing)
 predictions <- predict(modFit,newdata=testingImputed)
```

Now we would like to compare the in sample error and the out of sample error.
```{r SampleErrors , dependson='preditTestingDataFrame'}
# the in sample accuracy of the prediction model
fm <- modFit$finalModel
tail(fm$err.rate)
OOB_error<-fm$err.rate[nrow(fm$err.rate),1]

# out of sample accuracy of the predicted model
 outOfSampleError.accuracy <- sum(predictions == testingclasse)/length(predictions)
 outOfSampleError.accuracy

# out of sample error and percentage of out of sample error
 outOfSampleError <- 1 - outOfSampleError.accuracy
 outOfSampleError
```
The in OOB estimated sample error based on training with repeated cross validation is `r OOB_error` whereas the out of sample error based on the seperate test data frame is `r outOfSampleError` .


